Formalizing Stack Safety as a Security Property

### 2023-02-03 submission5/core.tex
abstract + intro

The term {\em stack safety} is associated with a
variety of compiler, run-time, and hardware mechanisms for protecting stack
memory. But, unlike ``the heap,'' the stack does not correspond to a single
high-level language concept. Rather, it corresponds to the fundamental
abstraction of functions, in all of the forms that they can take.
The protean nature of the functional abstraction makes stack safety
difficult to specify.

@@

The stack interacts with a large number of language features that are often excluded
from discussion of stack safety. Our properties support a system with
both caller- and callee-saved registers, arguments passed on the stack,
tail-call elimination, \ifexceptions exceptions \fi. They are modular by
design, in order to be further extensible to other features.

@@

Indeed, the call stack is an ancient~\cite{one06phrack} and
perennial~\cite{mitre-cwe,DBLP:conf/raid/VeendCB12,
  DBLP:conf/sp/SzekeresPWS13,
  DBLP:conf/sp/HuSACSL16,msrc-bluehat,chromium-security}
target for low-level attacks, frequently involving control-flow hijacking via corrupting
the return address, and memory corruption more generally.

%
%The most recent release of the authoritative CWE Top 25
%Most Dangerous Software Weaknesses~\cite{
%list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
%shows that general classes of vulnerabilities closely related to the stack
%% \rb{(though not exclusively)}
%consistently rank at and near the top of the list, e.g., \#1 Out-of-bounds Write, \#5 Out-of-bounds Read, and \#7 Use After Free.
%
%Moreover, many other types of faults involve some kind of control or data flow
%using data stored in the stack 
The variety in attacks on the stack is mirrored by the
software and hardware protections that aim to prevent them,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges22:TempsDesCerises},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18,Gollapudi+23}.

Security mechanisms can be brittle, successfully eliminating one attack while leaving room for
other, similar attacks. To avoid this, we seek formal properties of safe behavior that can be proven
or tested. Such properties become the specification against which enforcement can be validated;
even enforcement mechanisms that do not fulfill a property benefit from the ability to articulate
\emph{why} they fail.

@@

We show that we can extend our system to new features by first introducing a simple
example model, and then moving to a more realistic model with multiple new features:
callee-save registers, argument passing on the stack, and tailcall elimination, simply
by refining a few definitions in (1). This becomes the model that we validate.
In \cref{app:ptr}, we extend the model further with arbitrary sharing of memory-safe
pointers. Our criteria in (3) cover a total of five core requirements:
\(\wbcf\), caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)).
They are related to one another as seen in
\cref{fig:propertydiagram}: \(\clei\) subsumes \(\clrc\) and \(\clec\) subsumes
\(\clri\), but otherwise the properties are orthogonal and can be taken piecemeal
according to the needs of a system.

In particular, many enforcement techniques focus purely on
well-bracketed control flow. Stack canaries aim to prevent certain attacks on the return
address, and shadow stacks with protection (e.g. Return Address Defender \cite{Chiueh2001RAD})
to enforce it completely. Others combine this protection with some degree of memory protection,
chiefly focusing on integrity. Interestingly, Skorstengaard et al. \cite{SkorstengaardSTKJFP}
describe their ``local state encapsulation'' in terms of integrity, but it is equivalent
to the combination of \(\clri\) and \(\clrc\). The follow-up work by 
Georges et al. \cite{Georges22:TempsDesCerises} cites an early preprint of our work
as inspiration for extending their formal discussion to include confidentiality.
When checking if a preliminary version of these properties applied to
the Cerise calling convention, they noted that the Skostengaard et al. calling convention
did not enforce callee confidentiality, and made sure that their version would
\cite{Georges22:personalcommunication}, in addition to building it into their formalism.
This demonstrates the benefit of our choice to explicitly state properties in security
terms: specifying security mechanisms is hard, and when the specification takes the
form of a ``correct by construction'' machine, it is easy to neglect a non-obvious
security requirement.

@@

Our artifact contains Coq formalizations of our properties for illustrative purposes
as well as their testing implementation. There are no proofs---we use Coq primarily
for the QuickChick library and to ensure that the formalizations are unambiguous.
Proofs are future work.

@@

\paragraph{Organization}

@@

\newcommand{\freebox}[1][10px]{\tikz \filldraw[fill=blue] (0,0) rectangle (#1,#1);}
\newcommand{\pubbox}[1][10px]{\tikz \filldraw[fill=lightgray] (0,0) rectangle (#1,#1);}
\newcommand{\objbox}[1][10px]{\tikz \filldraw[fill=yellow] (0,0) rectangle (#1,#1);}
\newcommand{\sealbox}[1][10px]{\tikz \filldraw[fill=red] (0,0) rectangle (#1,#1);}

\newcommand{\textbox}[1][]{\tikz \filldraw[fill=lightgray] (0,0) rectangle (30px,30px) node[pos=.5]{#1};}
\newcommand{\numbox}[1][]{\tikz \filldraw[fill=lightgray] (0,0) rectangle (20px,20px) node[pos=.5]{#1};}
\newcommand{\leftbox}[1][]{\tikz \filldraw[fill=blue!20] (0,0) rectangle (20px,20px) node[pos=.5]{#1};}
\newcommand{\rightbox}[1][]{\tikz \filldraw[fill=orange!20] (0,0) rectangle (20px,20px) node[pos=.5]{#1};}

@@

      \}}
  \end{subfigure}
  \begin{subfigure}{\columnwidth}
    \begin{tabular}{r l | l}
      \labeledrow{0:}{addi sp,sp,-20}{\(\mathbf{alloc} ~ (-20,20)\)}
      \labeledrow{4:}{sd ra,12(sp)}{}
      \labeledrow{8:}{sw a0,8(sp)}{}
      \labeledrow{12:}{sw zero,4(sp)}{}
      \labeledrow{16:}{jal f,ra}{\(\mathbf{call} ~ \emplist \)}
      \labeledrow{20:}{sw a0,0(sp)}{}
      \labeledrow{24:}{lw a4,4(sp)}{}
      \labeledrow{28:}{li a5,42}{}
      \labeledrow{32:}{bne a4,a5,L1}{}
      \labeledrow{36:}{lw a0,8(sp)}{}
      \labeledrow{40:}{sw a0,out}{}
      \labeledrow{44:}{j L2:}{}
      \labeledrow{L1, 48:}{lw a0,0(sp)}{}
      \labeledrow{52:}{sw a0,out}{}
      \labeledrow{L2, 56:}{ld ra,12(sp)}{}
      \labeledrow{60:}{addi sp,sp,20}{\(\mathbf{dealloc} ~ (0,20)\)}
      \labeledrow{64:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
  \end{subfigure}
  \begin{subfigure}{\columnwidth}
    \center
    \vspace{\abovedisplayskip}
    \(\textnormal{\large \dots} \pubbox[30px] \!\! \stackrel{\textsc{\normalsize sp}}{\textbox[\tt res]}
    \stackrel{\textsc{\normalsize 4(sp)}}{\textbox[\tt sens]}
    \stackrel{\textsc{\normalsize 8(sp)}}{\textbox[\tt sec]}
    \stackrel{\textsc{\normalsize 12(sp)} \hfill}{\textbox[\(\mathtt{ra_1}\)]\textbox[\(\mathtt{ra_2}\)]}
    \textnormal{\large \dots}\)
  \end{subfigure}

>>>>

The term {\em stack safety} is associated with a
variety of compiler, run-time, and hardware mechanisms for protecting stack
memory. But, unlike ``the heap,'' the stack does not correspond to a single
high-level language concept. Rather, it corresponds to the fundamental
abstraction of functions, in all of the forms that they can take.
This protean nature of the functional abstraction makes stack safety
difficult to specify.

@@

The stack interacts with a large number of language features that are often excluded
from discussion of stack safety. Our properties support a system with
both caller- and callee-saved registers, arguments passed on the stack,
\ifexceptions tail-call elimination, and exceptions.
\else and tail-call elimination. \fi They are modular by
design, in order to be further extensible to other features.

@@

Indeed, the call stack is an ancient~\cite{phrack96:smashingthestack} and
perennial~\cite{mitre-cwe,DBLP:conf/raid/VeendCB12,
  DBLP:conf/sp/SzekeresPWS13,
  DBLP:conf/sp/HuSACSL16,msrc-bluehat,chromium-security}
target for low-level attacks, frequently involving control-flow hijacking via corrupting
the return address, and memory corruption more generally.
%
%The most recent release of the authoritative CWE Top 25
%Most Dangerous Software Weaknesses~\cite{
%list\footnote{\url{https://cwe.mitre.org/top25/archive/2022/2022_cwe_top25.html}}
%shows that general classes of vulnerabilities closely related to the stack
%% \rb{(though not exclusively)}
%consistently rank at and near the top of the list, e.g., \#1 Out-of-bounds Write, \#5 Out-of-bounds Read, and \#7 Use After Free.
%
%Moreover, many other types of faults involve some kind of control or data flow
%using data stored in the stack 
The variety in attacks on the stack is mirrored by the
software and hardware protections that aim to prevent them,
%
including stack canaries~\cite{Cowan+98},
bounds checking~\cite{NagarakatteZMZ09,NagarakatteZMZ10,DeviettiBMZ08},
split stacks~\cite{Kuznetsov+14},
shadow stacks~\cite{Dang+15,Shanbhogue+19},
capabilities~\cite{Woodruff+14,Chisnall+15,SkorstengaardLocal,SkorstengaardSTKJFP,Georges22:TempsDesCerises},
and hardware tagging~\cite{DBLP:conf/sp/RoesslerD18,Gollapudi+23}.

Security mechanisms can be brittle, successfully eliminating one attack while leaving room for
other, similar attacks. To avoid this \leo{game of whack-a-mole}, we seek formal properties of safe behavior that can be proven
or tested. Such properties become the specification against which enforcement can be validated;
even enforcement mechanisms that do not fulfill a property benefit from the ability to articulate
\emph{why} they fail.

@@

We show that we can extend our system to new features by first introducing a simple
example model, and then moving to a more realistic model with multiple new features:
callee-save registers, argument passing on the stack, and tailcall elimination, simply
by refining a few definitions in (1). This becomes the model that we validate.
In \cref{app:ptr}, we extend the model further with arbitrary sharing of memory-safe
pointers. Our criteria in (3) cover a total of five core requirements:
\(\wbcf\), caller integrity (\(\clri\)), caller confidentiality (\(\clrc\)),
callee integrity (\(\clei\)), and callee confidentiality (\(\clec\)).
They are related to one another as seen in
\cref{fig:propertydiagram}:\leo{is this supposed to be \cref{tab:props}?} \(\clei\) subsumes \(\clrc\) and \(\clec\) subsumes
\(\clri\), but otherwise the properties are orthogonal and can be taken piecemeal
according to the needs of a system.

In particular, many enforcement techniques focus purely on
well-bracketed control flow. Stack canaries aim to prevent certain attacks on the return
address, and shadow stacks with protection (e.g. Return Address Defender \cite{Chiueh2001RAD})
to enforce it completely. Others combine this protection with some degree of memory protection,
chiefly focusing on integrity. Interestingly, Skorstengaard et al. \cite{SkorstengaardSTKJFP}
describe their ``local state encapsulation'' in terms of integrity, but it is equivalent
to the combination of \(\clri\) and \(\clrc\). In fact, the follow-up work by 
Georges et al. \cite{Georges22:TempsDesCerises} was inspired by the properties presented in this paper to extend their formal discussion to include confidentiality.
When checking if a preliminary version of these properties applied to
the Cerise calling convention, they noted that the Skostengaard et al. calling convention
did not enforce callee confidentiality, and made sure that their version would
\cite{Georges22:personalcommunication}, in addition to building it into their formalism.
This demonstrates the benefit of our choice to explicitly state properties in security
terms: specifying security mechanisms is hard, and when the specification takes the
form of a ``correct by construction'' machine, it is easy to neglect a non-obvious
security requirement.

@@

Our artifact contains Coq formalizations of our properties for illustrative purposes
as well as their testing implementation. There are no proofs---we use Coq primarily
for the QuickChick library and to ensure that the formalizations are unambiguous.
Formal proofs are left as future work.

@@

\paragraph*{Organization}

@@

\newcommand{\freebox}[1][10pt]{\tikz \filldraw[fill=blue] (0,0) rectangle (#1,#1);}
\newcommand{\pubbox}[1][10pt]{\tikz \filldraw[fill=lightgray] (0,0) rectangle (#1,#1);}
\newcommand{\objbox}[1][10pt]{\tikz \filldraw[fill=yellow] (0,0) rectangle (#1,#1);}
\newcommand{\sealbox}[1][10pt]{\tikz \filldraw[fill=red] (0,0) rectangle (#1,#1);}

\newcommand{\textbox}[1][]{\tikz \filldraw[fill=lightgray] (0,0) rectangle (30pt,30pt) node[pos=.5]{#1};}
\newcommand{\numbox}[1][]{\tikz \filldraw[fill=lightgray] (0,0) rectangle (20pt,20pt) node[pos=.5]{#1};}
\newcommand{\leftbox}[1][]{\tikz \filldraw[fill=blue!20] (0,0) rectangle (20pt,20pt) node[pos=.5]{#1};}
\newcommand{\rightbox}[1][]{\tikz \filldraw[fill=orange!20] (0,0) rectangle (20pt,20pt) node[pos=.5]{#1};}

@@

      \}}
  \end{subfigure}
  \begin{subfigure}{\columnwidth}
    \begin{tabular}{r l | l}
      \labeledrow{0:}{addi sp,sp,-20}{\(\mathbf{alloc} ~ (-20,20)\)}
      \labeledrow{4:}{sd ra,12(sp)}{}
      \labeledrow{8:}{sw a0,8(sp)}{}
      \labeledrow{12:}{sw zero,4(sp)}{}
      \labeledrow{16:}{jal f,ra}{\(\mathbf{call} ~ \emplist \)}
      \labeledrow{20:}{sw a0,0(sp)}{}
      \labeledrow{24:}{lw a4,4(sp)}{}
      \labeledrow{28:}{li a5,42}{}
      \labeledrow{32:}{bne a4,a5,L1}{}
      \labeledrow{36:}{lw a0,8(sp)}{}
      \labeledrow{40:}{sw a0,out}{}
      \labeledrow{44:}{j L2:}{}
      \labeledrow{L1, 48:}{lw a0,0(sp)}{}
      \labeledrow{52:}{sw a0,out}{}
      \labeledrow{L2, 56:}{ld ra,12(sp)}{}
      \labeledrow{60:}{addi sp,sp,20}{\(\mathbf{dealloc} ~ (0,20)\)}
      \labeledrow{64:}{jalr ra}{\(\mathbf{return}\)}
    \end{tabular}
  \end{subfigure}
  \begin{subfigure}{\columnwidth}
    \center
    \vspace{\abovedisplayskip}
    \(\textnormal{\large \dots} \pubbox[30pt] \!\! \stackrel{\textsc{\normalsize sp}}{\textbox[\tt res]}
    \stackrel{\textsc{\normalsize 4(sp)}}{\textbox[\tt sens]}
    \stackrel{\textsc{\normalsize 8(sp)}}{\textbox[\tt sec]}
    \stackrel{\textsc{\normalsize 12(sp)} \hfill}{\textbox[\(\mathtt{ra_1}\)]\textbox[\(\mathtt{ra_2}\)]}
    \textnormal{\large \dots}\)
  \end{subfigure}
